### 2023-10-25 paper/paper-papaja.Rmd
minor revisions and prep for submission

  Materials for all experiments can be found at https://osf.io/znq3y/. Preregistration information for Experiments can be found at https://osf.io/c4w8s/ (Exp. 1), https://osf.io/czkdh/ (Exp. 2), and https://osf.io/5ewa6 (Exp. 3).

abstract: |
  Numerous psychological findings have shown that mere exposure to ideas makes those ideas seem more true, a finding commonly referred to as the “illusory truth” effect [e.g. @hasher.etal1977]. In the presence of pervasive misinformation, this basic feature of cognition may undermine the functioning of a democratic society [@pennycook.etal2018]. However, genuine beliefs do not only affect judgments about individual statements, they also imply other beliefs and drive decision-making. Here, we sought to examine whether mere exposure to statements produces genuine beliefs by examining whether people draw inferences from statements after mere exposure. Surprisingly, and in contrast to fluency or familiarity-based accounts of the illusory truth effect [e.g. @dechene.etal2010], we found that exposure to “premise” statements affected participants’ truth ratings for novel “implied” statements. This “illusory implication” effect suggests that the consequences of exposure to false statements reach further than previously thought and calls for a new mechanistic account of these effects.

@@

keywords          : "Illusory truth; Metacognition; Cognitive Psychology; Misinformation"
wordcount         : "4930"

@@

## Relevance Statement

Three pre-registered experiments examined whether mere exposure to ideas might lead them to be adopted as beliefs. It is well-established that mere exposure, such as in a true-or-false quiz, can lead people to later rate statements as more true (the “illusory truth” effect). If people's beliefs are affected by mere exposure, then the pervasiveness of misinformation would be even more harmful than currently recognized. However, leading theories attribute illusory truth effects to feelings of familiarity or fluency rather than changes in beliefs. These metacognitive heuristics are unlikely to produce further inferences or drive decision-making, as would beliefs. In contrast, we demonstrate that mere exposure to statements affected participants' responses to other novel statements (what we term an “illusory implication” effect), suggesting that mere exposure can lead to genuine belief. Our findings illustrate new dangers in exposure to misinformation, showing that mere exposure may have more wide-ranging influences than previously understood.

@@

There is something unsettling about the illusory truth effect and the lack of agency it implies over our own beliefs. Even more worrisome, there could be dire societal implications if merely being exposed to an idea causes people to adopt it as a belief: Pennycook and colleagues [-@pennycook.etal2018] argue that the illusory truth effect, combined with an environment of pervasive misinformation, has important implications for the functioning of democratic society. The illusory truth effect suggests that mere exposure to misinformation could have impacts that cannot be stopped by fact-checking labels [@pennycook.etal2018], nor effectively curbed by retractions or corrective information [c.f. @ecker.etal2011; @lewandowsky.etal2012].

@@

A handful of studies are suggestive in this regard: Arkes and collagues [-@arkes.etal1991] found that mere exposure to one set of statements also affected judgments about novel related statements on the same topic (China). However, these findings could also be explained by a sense of familiary engendered by the shared topic of old and new statements [@arkes.etal1991].

@@

@@

>>>>

  Materials, data, and analysis code for all experiments can be found at https://osf.io/znq3y/. Preregistration information for Experiments can be found at https://osf.io/c4w8s/ (Exp. 1), https://osf.io/czkdh/ (Exp. 2), and https://osf.io/5ewa6 (Exp. 3). Experiments 1 and 2 were first reported in a six-page paper presented at the _43rd Annual Meeting of the Cognitive Science Society_ and appearing in the conference's _Proceedings_. This non-archival conference paper included similar narrative motivations for the studies as presented here, with a more limited literature review. The narrative interpretations and discussion have changed in light of the findings of the novel third experiment and additional analyses reported here.

abstract: |
  Numerous psychological findings have shown that mere exposure to ideas makes those ideas seem more true, a finding commonly referred to as the “illusory truth” effect. This basic feature of cognition may hamper efforts to curb misinformation: critical media consumption and fact-checking seem poised to fail if mere exposure can persuade people to adopt ideas. Under predominant accounts of the illusory truth effect, initial exposure to a statement provides a metacognitive feeling of “fluency” or familiarity that, upon subsequent exposure, leads people to infer the statement is more likely to be true. However, genuine beliefs do not only affect truth judgments about individual statements, they also imply other beliefs and drive decision-making. Here, we conducted three pre-registered experiments to examine whether mere exposure to statements produces genuine beliefs by examining whether people draw inferences from statements after mere exposure. Surprisingly, and in contrast to fluency or familiarity-based accounts of the illusory truth effect, we found that exposure to “premise” statements affected participants’ truth ratings for novel “implied” statements. This “illusory implication” effect suggests that the consequences of exposure to mis- and disinformation reach further than previously thought and calls for a new mechanistic account of these effects.

@@

keywords          : "Illusory truth; Metacognition; Cognitive Psychology; Misinformation"
wordcount         : "5099"

@@

## Public Significance Statement

It is well-established that mere exposure to statements, such as in a true-or-false quiz, can lead people to later rate those same statements as more true (the “illusory truth” effect). On most accounts, the illusory truth effect is thought to be caused by fluency or familiarity with a specific statement, and therefore should not more widely affect people's beliefs or decisions in most cases. In contrast to these prior accounts, however, we found that mere exposure to statements affected participants' responses to other novel statements (what we term an “illusory implication” effect), suggesting that mere exposure can instead lead to genuine belief. Our findings are further cause for concern about the spread and influence of misinformation, as they indicate that any exposure to misinformation may cause harm by inducing false beliefs.

@@

There is something unsettling about the illusory truth effect and the lack of agency it implies over our own beliefs. Even more worrisome, there could be important societal implications if merely being exposed to an idea causes people to adopt it as a belief: Pennycook and colleagues [-@pennycook.etal2018] argue that the illusory truth effect, combined with an environment of pervasive misinformation, has important consequences for the functioning of democratic society. The illusory truth effect suggests that mere exposure to misinformation could have impacts that cannot be stopped by fact-checking labels [@pennycook.etal2018], nor effectively curbed by retractions or corrective information [c.f. @ecker.etal2011; @lewandowsky.etal2012].

@@

A handful of studies are suggestive in this regard: Unkelbach and Rom [-@unkelbach.rom2017] found that mere exposure to one statement could increase endorsements of implied statements. For instance, being more likely to endorse the statement "Most accidents occur close to weekends" after exposure to the statement "Most accidents occur on Mondays". Similarly, Arkes and colleagues [-@arkes.etal1991] found that mere exposure to one set of statements also affected judgments about novel related statements on the same topic (China). However, in both cases these findings can also be easily explained by fluency or familiarity engendered by the similarity between the presented and tested statements [@arkes.etal1991; @unkelbach.rom2017].

@@

# Transparency and Openness

All experiments were preregistered prior to data collection and all materials, data, and analysis code have been made publicly available on the Open Science Foundation website (https://osf.io/znq3y/). We report our approach to determining sample sizes, all data exclusions, all manipulations, and all measures in each study. Where appropriate, we explicitly differentiate between preregistered analyses and non-preregistered analyses in our results. All statistical models were fit using the `brms` R package [@burkner2017]. Our study was reviewed and approved by the Arizona State University Institutional Review Board under project identifier `STUDY00013322`. Participants provided informed consent before beginning the study.

@@

## Constraints on Generality

The studies reported here targeted U.S. adults through an online survey provider (CloudResearch). Study materials were designed accordingly and may produce different results if tested on populations with different backgrounds and prior knowledge.
